What is an Ancestral Recombination Graph?

### 2022-03-01 paper.tex
Update refs

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2021unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies,
leading to basic errors in statements about what different inference methods
produce as output (see the XXX section below), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

% TODO not clear this belongs in the discussion, but we'll definitely
% want to make some of these points, even if they are being repeated
% from earlier section.
It is difficult to imagine a practical calculation performed on an
ARG that does not involve either the sequential generation of trees
along the genome or the resolved ancestry segments.
Knowing the resolved segments in an ancestry resolved GARG allows
us to perform many calculations without explicitly generating
the trees, and lets us work with the ancestry on a genome-by-genome
basis. We give three examples here. Firstly, generating mutations
on edges can be performed very quickly on an ancestry resolved
GARG, since we only need to examine $O(|E|)$ edges rather than
$O(nt)$ tree branches (assuming $n$ samples and $t$ marginal trees)
~\citep{baumdicker2021efficient}. If the ancestry were not resolved,
we would generate a large number of mutations on areas of the edge that
are non-ancestral. Secondly, computing the likelihood of a given
ARG realisation under the coalescent with recombination model
requires that we essentially recapitulate the state of Hudson's
algorithm (see XXX section) at each node in
the graph~\citep{baumdicker2021efficient}. Having resolved ancestry
substantially simplifies this process. Thirdly, computing
[FIXME describe what we did in tsdate]~\citep{wohns2021unified}.

Sequentially generating the marginal trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified GARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,kelleher2018efficient},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2021unified}
and likely many more.

>>>>

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies,
leading to basic errors in statements about what different inference methods
produce as output (see the XXX section below), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

@@

% TODO not clear this belongs in the discussion, but we'll definitely
% want to make some of these points, even if they are being repeated
% from earlier section.
It is difficult to imagine a practical calculation performed on an
ARG that does not involve either the sequential generation of trees
along the genome or the resolved ancestry segments.
Knowing the resolved segments in an ancestry resolved GARG allows
us to perform many calculations without explicitly generating
the trees, and lets us work with the ancestry on a genome-by-genome
basis. We give three examples here. Firstly, generating mutations
on edges can be performed very quickly on an ancestry resolved
GARG, since we only need to examine $O(|E|)$ edges rather than
$O(nt)$ tree branches (assuming $n$ samples and $t$ marginal trees)
~\citep{baumdicker2021efficient}. If the ancestry were not resolved,
we would generate a large number of mutations on areas of the edge that
are non-ancestral. Secondly, computing the likelihood of a given
ARG realisation under the coalescent with recombination model
requires that we essentially recapitulate the state of Hudson's
algorithm (see XXX section) at each node in
the graph~\citep{baumdicker2021efficient}. Having resolved ancestry
substantially simplifies this process. Thirdly, computing
[FIXME describe what we did in tsdate]~\citep{wohns2022unified}.

Sequentially generating the marginal trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified GARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,kelleher2018efficient},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2022unified}
and likely many more.
